---
title: "Statistical Learning Project"
author: "Marlon Helbing, Daniele Virzì"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This project is a collaborative effort between two students, **Marlon Helbing** and **Daniele Virzì**. It is an academic project that will be graded based on the quality and depth of our analysis. The project aims to apply the concepts and techniques learned in the **Statistical Learning Course** to a real-world dataset.

## Project goals

The main objectives of this project are:

1.  **Exploratory Data Analysis (EDA)**: We aim to uncover patterns and trends in the data that could provide insights into what makes a song popular on ***Spotify***.

2.  **Regression Model**: This model aims to predict a song's popularity, in terms of streams, based on various features present in the dataset.

## Dataset

- [**Most Streamed Spotify Songs 2023**](https://www.kaggle.com/datasets/nelgiriyewithana/top-spotify-songs-2023): This dataset, sourced from Kaggle, is a comprehensive collection of the most streamed songs of 2023 as listed on Spotify. It was chosen for our project due to our shared love for music and our domain knowledge in this area. It provides insights into each song’s attributes, popularity, and presence on various music platforms. This makes it an excellent resource for our project, as it allows us to explore a wide range of factors that could potentially influence a song’s popularity.

The dataset contains the following columns:

-   **`track_name`**: *Name of the song*.

-   **`artist(s)_name`**: *Name of the artist(s) of the song*.

-   **`artist_count`**: *Number of artists contributing to the song*.

-   **`released_year`**: *Year when the song was released*.

-   **`released_month`**: *Month when the song was released*.

-   **`released_day`**: *Day of the month when the song was released*.

-   **`in_spotify_playlists`**: *Number of Spotify playlists the song is included in*.

-   **`in_spotify_charts`**: *Presence and rank of the song on Spotify charts*.

-   **`streams`**: *Total number of streams on Spotify*.

-   **`in_apple_playlists`**: *Number of Apple Music playlists the song is included in*.

-   **`in_apple_charts`**: *Presence and rank of the song on Apple Music charts*.

-   **`in_deezer_playlists`**: *Number of Deezer playlists the song is included in*.

-   **`in_deezer_charts`**: *Presence and rank of the song on Deezer charts*.

-   **`in_shazam_charts`**: *Presence and rank of the song on Shazam charts*.

-   **`bpm`**: *Beats per minute, a measure of song tempo*.

-   **`key`**: *Key of the song*.

-   **`mode`**: *Mode of the song (major or minor)*.

-   **`danceability_%`**: *Percentage indicating how suitable the song is for dancing*.

-   **`valence_%`**: *Positivity of the song's musical content*.

-   **`energy_%`**: *Perceived energy level of the song*.

-   **`acousticness_%`**: *Amount of acoustic sound in the song*.

-   **`instrumentalness_%`**: *Amount of instrumental content in the song*.

-   **`liveness_%`**: *Presence of live performance elements*.

-   **`speechiness_%`**: *Amount of spoken words in the song*.

# Data Loading

As previously stated, the purpose of this project is to assess the knowledge we have gained from the **Statistical Learning Course**. Because of this, we were only permitted to utilize the models and techniques covered in the course for our project work; we were not permitted to use any data analysis techniques that were not covered in class or any `Tidyverse` R-packages, like `ggplot` or `ggplot2`.

## Packages required

```{r imports, echo=FALSE}}
library(corrplot)
```

## Loading the dataset

```{r data, echo=FALSE}
Spotify <- read.csv("spotify-2023.csv")
```

# Data Cleaning

In this section, we had performed some data cleaning tasks to ensure that the dataset is ready for analysis. This includes checking and handling appropriately missing values and duplicates.

## Checking the structure of the data

```{r explore}
# Check the first 5 rows of the dataset in order to understand the structure of the data
head(Spotify)
```

## Handling missing values

```{r missing}
# Check how many missing values there are in the dataset and where they are
missing_values <- sum(is.na(Spotify))

# Remove all rows where there is atleast one missing value in some column
Spotify <- na.omit(Spotify)
```

## Handling duplicates

```{r duplicates}
# Check how many duplicates there are in the dataset
duplicates <- sum(duplicated(Spotify))

# Remove all duplicates from the dataset
Spotify <- unique(Spotify)
```

# Data Preprocessing

In order to perform the analysis, we need to preprocess the data. This includes changing the datatype of some columns, removing unnecessary columns, and handling categorical values.

```{r structure}
# Check the structure of the dataset in order to understand the datatypes
str(Spotify)
```

## Date datatype

We had the date of release given in three different columns (day,month,year) as strings and we wanted to change this into one column and save it as datatype date. Typical format for date is '*2022-06-22*'. As a way to achieve this result, we used `paste()`with the parameter `sep = '-'` to aggregate the three columns into a new column , and then we converted the column datatype using `as.Date()` with parameter `format = "%Y-%m-%d"`. We dropped the original three columns since they were no longer relevant.

```{r date}
# First we create a new column and store the date in the wanted format 
Spotify$released_date <- paste(Spotify$released_year,
                               Spotify$released_month,
                               Spotify$released_day,
                               sep = '-')

# We give it the wanted datatype of DATE
Spotify$released_date <- as.Date(Spotify$released_date,
                                 format = "%Y-%m-%d")

# We can directly drop our other 3 columns storing the dates
Spotify <- subset(Spotify, select = -c(released_year,
                                       released_month,
                                       released_day))
```

## Categorical datatype

We have noticed that tha columns **`key`** and **`mode`** were *categorical* and we wanted to convert them into *factors* using `as.factor()`*.*

```{r factor}
# Change columns with categorical values into the right format (i.e. factor)
Spotify$key <- as.factor(Spotify$key)
Spotify$mode <- as.factor(Spotify$mode)
```

Moreover, the columns **`in_deezer_playlists`**, and **`in_shazam_charts`** , unlike the other playlists and charts features, were *categorical* so we have converted them into *integer* using `as.integer()`.

```{r integer}
# Change columns with categorical values into the right format (i.e. integer)
Spotify$in_deezer_playlists <- as.integer(Spotify$in_deezer_playlists)
Spotify$in_shazam_charts <- as.integer(Spotify$in_shazam_charts)
```

# Data Summary

```{r summary}
# Check the summary of the dataset in order to understand the distribution of the data
summary(Spotify)
```
